An accelerometer-derived ballistocardiogram method for detecting heartrates in free-ranging marine mammals

### 2021-10-27 analysis/paper/paper.Rmd
Incorporate co-author feedback

---
title: "An accelerometer-derived ballistocardiogram method for detecting heartrates in free-ranging marine mammals"
author:
  - Max F. Czapanskiy:
      email: maxczap@stanford.edu
      institute: [Hopkins]
      correspondence: true
  - Paul J. Ponganis:
      institute: [Scripps]
      correspondence: false
  - James A. Fahlbusch:
      institute: [Hopkins]
      correspondence: false
  - Jeremy A. Goldbogen:
      institute: [Hopkins]
      correspondence: false
institute:
  - Hopkins: Hopkins Marine Station, Department of Biology, Stanford University, USA
  - Scripps: Scripps Institution of Oceanography, University of California San Diego, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx"
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: [zotero.bib,references.bib]
csl: "../templates/the-journal-of-experimental-biology.csl"
abstract: |
  Physio-logging methods, which use animal-borne devices to record physiological variables, are entering a new era driven by advances in sensor development. However, existing datasets collected with traditional bio-loggers, such as accelerometers, still contain untapped eco-physiological information. Here we present a computational method for extracting heartrate from high-resolution accelerometer data using a ballistocardiogram. We validated our method with simultaneous accelerometer-electrocardiogram tag deployments in a controlled setting on a killer whale (*Orcinus orca*) and demonstrate the method recovers previously observed cardiovascular patterns in a blue whale (*Balaenoptera musculus*), including the magnitude of apneic bradycardia and progressive relaxation of bradycardia during dives. Our ballistocardiogram method may be applied to mine heart rates from previously collected accelerometery and expand our understanding of comparative cardiovascular physiology.
keywords: |
  ballistocardiogram; diving physiology; cardiovascular physiology; accelerometer; bio-logging; marine mammal
---

@@

Recent advances in physio-logging (recording physiological variables using animal-borne devices) have largely been driven by new developments in sensor technology [@hawkesIntroductionThemeIssue2021]. For example, new physio-logging tags can detect regional changes in blood flow by incorporating functional near-infrared spectroscopy sensors [@mcknightShiningNewLight2021]. However, legacy tags with inertial measurement unit (IMU) sensors - accelerometers, magnetometers, and gyroscopes - can also measure important physiological variables under certain conditions. In one case, accelerometers revealed the elevated respiratory rate of an emperor penguin (*Aptenodytes forsteri*) following a 27.6 minute dive, the longest recorded for this species [@satoStrokeRatesDiving2011]. While physio-logging tags with cutting-edge biomedical technologies push the boundaries of physiological field research, simpler IMU tags have fewer logistical constraints and provide access to more species and larger sample sizes. This is particularly important for species that cannot be restrained or studied in captivity. For example, of the sixteen species of baleen whales (Mysticeti), heart rate has only been recorded with an electrocardiogram tag on one blue whale (*Balaenoptera musculus*) individual [@goldbogenExtremeBradycardiaTachycardia2019]. Conversely, IMU tags have been deployed on hundreds of individuals of nearly every species in the clade for the last twenty years [@nowacekBuoyantBalaenidsUps2001] . These existing datasets (and future IMU tag deployments) could hold valuable physiological information, but we lack proper computational methods for mining them.

The ballistocardiogram (BCG) has potential applications to using accelerometers as heartrate monitors. Ballistocardiography is a noninvasive method for measuring cardiac function based on the ballistic forces involved in the heart ejecting blood into the major vessels. The BCG originated as a clinical tool in the first half of the 20th century [@starrStudiesEstimationCardiac1939], but was largely superseded by electro- and phonocardiography. However the medical community has recently returned to ballistocardiography as a potential means of passive monitoring of heart function in at-risk populations [@giovangrandi2011ballistocardiography], which has led to substantial progress in signal processing methodology for generating interpretable BCGs [@sadekBallistocardiogramSignalProcessing2019]. While the BCG is a three-dimensional phenomenon, it is strongest in the longitudinal (anterior-posterior) axis [@inanBallistocardiographySeismocardiographyReview2015]. Along this axis, the waveform is composed of multiple peaks and valleys; most prominent of these are the IJK complex, which occurs during systole [@pinheiroTheoryDevelopmentsUnobtrusive2010]. The BCG J wave is analagous to the ECG R wave in that it is the most robust feature in the waveform and typically used for detecting heart beats.

Here we present a method for generating a BCG from bio-logger accelerometry. We validated our method with a simultaneously recorded electrocardiogram (ECG) on a captive killer whale (*Orcinus orca*) and applied it to detect heartrate in a blue whale. The relative orientation of the tag on the body is uncertain in cetacean bio-logging [@johnsonDigitalAcousticRecording2003], so in addition to a one-dimensional BCG based solely on longitudinal acceleration, we also generated a three-dimensional BCG, which we expected would be more robust in a field setting. Specifically, we tested three hypotheses to validate our method. First, a one-dimensional BCG would, in a controlled setting, produce statistically equivalent instantaneous heartrates as an ECG. Second, a three-dimensional BCG would, in a field setting, produce a more robust signal than a one-dimensional BCG. Third, BCG-derived heartrates would increase from the start to end of dives, consistent with the marine mammal cardiophysiological dive response of progressive bradycardia relaxation [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014].

@@

* A captive killer whale was double-tagged with a CATS unit and ECG recorder.
* The CATS unit was placed behind the left pectoral flipper and the ECG recorder on the ventral center line. 
* Trainers instructed the whale to hold a submerged position.
* A blue whale was tagged with a CATS unit (originally published in [@goldbogenWhyWhalesAre2019]).
* The tag slid behind the right pectoral flipper, where it remained overnight.
* While tagged, the blue whale engaged in apparent resting behavior. We manually labeled motionless periods. Specifically, we looked for regions with low amplitude y-axis gyroscope signal, since fluke-strokes are readily apparent in this signal [@goughScalingSwimmingPerformance2019]
* CATS tags programmed to collect 400 Hz accelerometer, 50 Hz magnetometer, 50 Hz gyroscope, and 10 Hz pressure. Data processed using CATS Toolbox [@cadeToolsIntegratingInertial2021].
* Relevant permits (IACUC, NMFS, ???)

@@

The BCG waveform is three dimensional, but strongest in the longitudinal axis [@inanBallistocardiographySeismocardiographyReview2015]. We tested 1d and 3d metrics for identifying heartbeats in acceleration data based on the methods of [@leePhysiologicalSignalMonitoring2016].

For windowed operations, we used 0.5 s for killer whale data and 2.0 s for blue whale data.

@@

1. Remove noise and de-trend the acceleration signal with a 5th order Butterworth band-pass filter (killer whale: [1-25Hz], blue whale: [1-10Hz]) using package `signal` [@R-signal].
2. Enhance the IJK complex by differentiating acceleration using a 4th order Savitzky-Golay filter using package `signal`. Differentiation exaggerates impulses like the J wave.
3. Further enhance the peaks by calculating the Shannon entropy ($H_i=-\sum_{k} |a_{ik}| \times ln(|a_{ik}|)$, where $k$ is the acceleration axis). Additionally, the Shannon entropy is strictly positive, which facilitates peak detection. In the 1d case, $k$ is surge only.
4. Remove noise by applying a triangular moving average smoother.
5. Extract peaks and heuristically remove noisy peaks (see supplemental).

@@

We tested whether the 3d BCG was more robust than 1d BCG in field data by comparing the signal-to-noise ratios. For both BCGs, we calculated the power spectral density using package `psd` [@Barbour2014]. Previously recorded blue whale apneic heart rate was 4-8 bpm [@goldbogenExtremeBradycardiaTachycardia2019], so we quantified *signal* as the integration of the power spectral density curve from 4-8 bpm and *noise* as the integrated remainder, up to 60 bpm.

We also tested whether BCG-derived instantaneous heart rates exhibited bradycardia relaxation over the course of dives, consistent with diving physiology patterns in marine mammals [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014]. We assigned dive start and end times when the whale swam deeper than 2 m, retaining dives that exceeded 10 m depth and 5 minutes duration. Dive times were normalized from 0 (start of dive) to 1 (end of dive). We regressed instantaneous heart rate against normalized dive time using robust Theil-Sen regression (to account for heteroscedascity) with package `RobustLinearReg` [@R-RobustLinearReg; @Sen-1968; @theilRankInvariantMethodLinear1992] and tested whether the slope was greater than 0.

@@

The data and code used in this analysis were packaged as a research compendium using package `rrtools` [@marwickPackagingDataAnalytical2018; @rrtools2019]. The research compendium was written as an R package so other researchers can read, run, and modify the methods described here.

@@

The ECG and BCG yielded nearly identical heart rate estimations (Fig. \@ref(fig:oo-bcg-ecg)). We collected 14 s of simultaneous ECG and BCG data during a motionless, submerged breath hold. BCG-derived instantaneous heart rates were within `r sprintf("%0.1f%% \u00B1 %0.1f%%", mean_bpm_diff * 100, sd_bpm_diff * 100)` of the ECG-derived rates (mean &#177; standard deviation). The ordinary least squares regression of BCG heartrates on ECG heartrates yielded a slope of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[2], bpm_stderr[2])` and intercept of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[1], bpm_stderr[1])` (mean &#177; standard error), which were not significantly different from the hypothesized 1 and 0, respectively.

@@

```{r motionless}
bw_elg <- bw180905_53_elg %>% 
  mutate(depth_min = map2_dbl(start, stop, function(t1, t2) {
    bw180905_53_10hz %>% 
      filter(between(dt, t1, t2)) %>% 
      pull(depth) %>% 
      min()
  }),
  depth_max = map2_dbl(start, stop, function(t1, t2) {
    bw180905_53_10hz %>% 
      filter(between(dt, t1, t2)) %>% 
      pull(depth) %>% 
      max()
  })
  )
n_motionless <- nrow(bw_elg)
motionless_minutes <- as.numeric(sum(bw_elg$stop - bw_elg$start), unit = "mins")
```

@@

The 3d BCG (Fig. \@ref(fig:bw-bcg)) produced a more robust signal than the surge-only 1d BCG. The signal-to-noise ratio was 2.00 for the triaxial BCG, compared to 0.17 for the surge-only BCG (Fig. \@ref(fig:psd-plot)).

@@

```{r brady-relax-model}
bw10hz <- bw180905_53_10hz %>%
  mutate(
    dive_id = split_dives(dt, depth,
                          surface = 2, min_depth = 10, min_dur = 5 * 60),
    dive_norm = normalize_dives(dive_id)
  )
bw_400hz$dive_norm <- approx(bw10hz$dt, bw10hz$dive_norm, bw_400hz$dt)$y

@@

3d BCG-derived heart rates exhibited a relaxation of bradycardia over the course of dives. Average heart rate increased from `r format(bpm_pred$bcg_bpm[1], digits = 2)` bpm at the start of dives to `r format(bpm_pred$bcg_bpm[nrow(bpm_pred)], digits = 2)` bpm at the end of dives (Theil-Sen regression, $p < 10^{-10}$) (Fig \@ref(fig:brady-relax)).

@@

Here we presented a ballistocardiogram method for detecting resting apneic heartrate in cetaceans using accelerometers. We validated the method in a controlled setting with simultaneous ECG and in a field setting by confirming expected physiological patterns. As accelerometer tags have been deployed on many cetacean species for multiple decades, this method may be applied to mine existing datasets and better understand how heartrate scales with body size and other biological factors. Even as the field of physio-logging progresses with new hardware innovations, this method demonstrates that computational advances can still derive new insights from traditional sensors.

@@

* Everyone who helped collect and process the blue whale data.
* The Sea World trainers.
* Anna Krystalli, Ben Marwick, Karthik Ram, Nicholas Tierney, and other members of the open science R community for developing tools and educational resources that facilitate open science practices.

@@

```{r oo-bcg-ecg, fig.cap="The ECG (A) and BCG (E) produced nearly identical heart beat predictions (red and blue points, respectively). B-D display the intermediate steps in the BCG signal processing procedure. B: surge after filtering, C: after differencing, D: Shannon entropy. Y-axis labeling follows Lee et al. (2016 Sensors) and y-axis values were excluded because the filtering process introduces magnitude distortion; only the relative shape of the signal is relevant to the analysis."}
t <- theme_classic() +
  theme(axis.title.x = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_blank())

@@

# B: Filtered surge
pB <- ggplot(oo_400hz, aes(dt, surge_filt)) +
  geom_line(size = 0.2) +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = bquote(f[HR])) +
  t +
  theme(axis.text.y = element_blank())

# C: Surge diff
pC <- ggplot(oo_400hz, aes(dt, surge_diff)) +
  geom_line(size = 0.2) +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = "dBCG") +
  t +
  theme(axis.text.y = element_blank())

# D: Shannon entropy
pD <- ggplot(oo_400hz, aes(dt, surge_se)) +
  geom_line(size = 0.2) +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = "seBCG") +
  t +
  theme(axis.text.y = element_blank())

# E: Smoothed
pE <- ggplot(oo_400hz, aes(dt, surge_smooth)) +
  geom_line(size = 0.2) +
  geom_point(data = filter(oo_400hz, bcg_beat), color = "blue") +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = "Y") +
  t +
  theme(axis.text.y = element_blank())

cowplot::plot_grid(pA, pB, pC, pD, pE,
                   align = "v", 
                   axis = "lr",
                   ncol = 1, 
                   labels = "AUTO")
```

@@

```{r bw-bcg, fig.cap="Example of signal processing for 3d BCG. A: filtered triaxial acceleration, B: after differencing, C: Shannon entropy, D: after smoothing. Identified heart beats in blue. Y-axis labeling follows Lee et al. (2016 Sensors) and y-axis values were excluded because the filtering process introduces magnitude distortion; only the relative shape of the signal is relevant to the analysis."}
bw_bcg_plot <- function(rid) {
  bcg_sample <- filter(bw_400hz, region_id == rid)

  t <- theme_classic() +
  theme(axis.title.x = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_blank())

  # A: Filtered surge
  pA <- bcg_sample %>% 
    rename(Surge = surge_filt, Sway = sway_filt, Heave = heave_filt) %>% 
    pivot_longer(Surge:Heave, names_to = "axis", values_to = "acc") %>% 
    mutate(axis = factor(axis, levels = c("Surge", "Sway", "Heave"))) %>% 
    ggplot(aes(dt, acc)) +
    geom_line(size = 0.2) +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    scale_y_continuous(n.breaks = 3) +
    facet_grid(rows = vars(axis)) +
    labs(y = bquote(f[HR])) +
    t +
    theme(axis.text.y = element_blank())

  # B: Jerk vectors
  pB <- bcg_sample %>% 
    mutate(jerk = set_names(as_tibble(jerk), c("Surge", "Sway", "Heave"))) %>% 
    unpack(jerk) %>% 
    pivot_longer(Surge:Heave, names_to = "axis", values_to = "jerk") %>% 
    mutate(axis = factor(axis, levels = c("Surge", "Sway", "Heave"))) %>% 
    ggplot(aes(dt, jerk)) +
    geom_line(size = 0.2) +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    scale_y_continuous(n.breaks = 3) +
    facet_grid(rows = vars(axis)) +
    labs(y = "dBCG") +
    t +
    theme(axis.text.y = element_blank())

>>>>

---
title: "An accelerometer-derived ballistocardiogram method for detecting heartrates in free-ranging marine mammals"
author:
  - Max F. Czapanskiy:
      email: maxczap@stanford.edu
      institute: [Hopkins]
      correspondence: true
  - Paul J. Ponganis:
      institute: [Scripps]
      correspondence: false
  - James A. Fahlbusch:
      institute: [Hopkins]
      correspondence: false
  - T. L. Schmitt:
      institute: [SeaWorld]
      correspondence: false
  - Jeremy A. Goldbogen:
      institute: [Hopkins]
      correspondence: false
institute:
  - Hopkins: Hopkins Marine Station, Department of Biology, Stanford University, USA
  - Scripps: Scripps Institution of Oceanography, University of California San Diego, USA
  - SeaWorld: Veterinary Services, SeaWorld of California, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx"
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: [zotero.bib,references.bib]
csl: "../templates/the-journal-of-experimental-biology.csl"
abstract: |
  Physio-logging methods, which use animal-borne devices to record physiological variables, are entering a new era driven by advances in sensor development. However, existing datasets collected with traditional bio-loggers, such as accelerometers, still contain untapped eco-physiological information. Here we present a computational method for extracting heartrate from high-resolution accelerometer data using a ballistocardiogram. We validated our method with simultaneous accelerometer-electrocardiogram tag deployments in a controlled setting on a killer whale (*Orcinus orca*) and demonstrate the method recovers previously observed cardiovascular patterns in a blue whale (*Balaenoptera musculus*), including the magnitude of apneic bradycardia and progressive relaxation of bradycardia during dives. Our ballistocardiogram method may be applied to mine heart rates from previously collected accelerometery and expand our understanding of comparative cardiovascular physiology.
keywords: |
  ballistocardiogram; diving physiology; cardiovascular physiology; accelerometer; bio-logging; marine mammal
---

@@

Recent advances in physio-logging (recording physiological variables using animal-borne devices) have largely been driven by new developments in sensor technology [@hawkesIntroductionThemeIssue2021]. For example, new physio-logging tags can detect regional changes in blood flow by incorporating functional near-infrared spectroscopy sensors [@mcknightShiningNewLight2021]. However, traditional inertial measurement unit (IMU) tags equipped with accelerometers and other inertial sensors can also measure important physiological and related variables. Through careful inspection and analysis of high-resolution acceleration, scientists have measured elevated respiration rates following record-breaking dives [@satoStrokeRatesDiving2011], near-continuous feeding by small cetaceans [@wisniewskaUltraHighForagingRates2016], social interactions between large cetaceans  [@goldbogenUsingAccelerometersDetermine2014], and important biomechanical variables including movement speed [@cadeDeterminingForwardSpeed2018]. While physio-logging tags with cutting-edge biomedical technologies push the boundaries of physiological field research, simpler IMU tags have fewer logistical constraints and provide access to more species and larger sample sizes. This is particularly important for species that cannot be restrained or studied in managed care. For example, of the sixteen species of baleen whales (Mysticeti), heart rate has only been recorded with an electrocardiogram tag in the wild for one blue whale (_Balaenoptera musculus_) [@goldbogenExtremeBradycardiaTachycardia2019; but see @ponganisHeartRateElectrocardiogram1999]. Conversely, IMU tags have been deployed on hundreds of individuals of nearly every species in the clade for the last twenty years [@nowacekBuoyantBalaenidsUps2001]. These existing datasets (and future IMU tag deployments) could hold additional valuable physiological information, awaiting proper computational methods for mining them.

The ballistocardiogram (BCG) has potential applications to using accelerometers as heartrate monitors in both the wild and in managed care. Ballistocardiography is a noninvasive method for measuring cardiac function based on the ballistic forces involved in the heart ejecting blood into the major vessels. The BCG originated as a clinical tool in the first half of the 20th century [@starrStudiesEstimationCardiac1939], but was largely superseded by electro- and echocardiography. However the medical community has recently returned to ballistocardiography as a potential means of passive monitoring of heart function in at-risk populations [@giovangrandi2011ballistocardiography], which has led to substantial progress in signal processing methodology for generating interpretable BCGs [@sadekBallistocardiogramSignalProcessing2019]. While the BCG is a three-dimensional phenomenon, it is strongest in the anterior-posterior axis [@inanBallistocardiographySeismocardiographyReview2015]. Along this axis, the waveform is composed of multiple peaks and valleys; most prominent of these are the IJK complex, which progressively occurs during systole [@pinheiroTheoryDevelopmentsUnobtrusive2010]. The BCG J wave is the most robust feature in the waveform and typically used for detecting heart beats.

Here we present a method for generating a BCG from bio-logger accelerometry. We validated our method with a simultaneously recorded electrocardiogram (ECG) on an adult killer whale in managed care (_Orcinus orca_) and applied it to detect heartrate in a blue whale. The relative orientation of the tag on the body is uncertain in cetacean bio-logging in the wild [@johnsonDigitalAcousticRecording2003], so in addition to a one-dimensional BCG based solely on anterior-posterior acceleration (surge), we also generated a three-dimensional BCG, which we expected would be more robust in a field setting. Specifically, we tested three hypotheses to validate our method. First, a one-dimensional BCG would, in a controlled setting, produce statistically equivalent instantaneous heartrates as an ECG. Second, a three-dimensional BCG would, in a field setting, produce a more robust signal than a one-dimensional BCG. Third, BCG-derived heartrates would increase during the bottom and latter phases of dives, consistent with the progressive increase in heartrate routinely observed before and during ascent [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014].

@@

**Killer whale**

A killer whale in managed care at the SeaWorld, San Diego, CA marine facilities was double-tagged with Customized Animal Tracking Solutions IMU (CATS, www.cats.is) and electrocardiogram (ECG) tags on August 16, 2021 as part of clinical animal cardiac evaluations under the SeaWorld display permit. The CATS tag recorded acceleration at 400 Hz, magnetometer and gyroscope at 50 Hz, pressure at 10 Hz, and video at 30 fps. All sensors were rotated from the tag's frame of reference to that of the whale using MATLAB (MathWorks, Inc., v2020b) tools for processing CATS data [@cadeToolsIntegratingInertial2021]. The ECG tag hardware and data processing followed the methods in [@bickettHeartRatesHeart2019]. Briefly, we recorded ECG at 100 Hz and identified individual heart beats from visually verified R-waves using a customized peak detection program (K. Ponganis; Origin 2017, OriginLab Co., Northampton, MA). ECG and IMU were recorded during a submerged breath hold.

**Blue whale**

A 24.5 m blue whale was tagged with a CATS IMU tag on September 5, 2018 in Monterey Bay, CA under permits MBNMS-MULTI-2017-007, NMFS 21678, and Stanford University IACUC 30123 (previously published in [@goughScalingSwimmingPerformance2019]). Tag configuration and data processing followed the same procedure as the killer whale, with one addition. The 400 Hz acceleration data was used for ballistocardiography (see section **Signal processing**), but we also downsampled the multi-sensor data to 10 Hz for movement analysis using the MATLAB CATS tools. Dynamic body movements produce an acceleration signal that masks the ballistocardiogram, so we limited our analyses to manually labeled motionless periods. Motionless periods were identified during or near the bottom phase of dives between fluke strokes. Strokes were detected from visual examination of the rotational velocity around the lateral axis recorded by gyroscope [_sensu_ @goughScalingSwimmingPerformance2019].

@@

The BCG waveform is three dimensional, but strongest in the anterior-posterior axis [@inanBallistocardiographySeismocardiographyReview2015]. We tested 1d and 3d metrics for identifying heartbeats in acceleration data based on the methods of [@leePhysiologicalSignalMonitoring2016]. For windowed operations, we used 0.5 s for killer whale data and 2.0 s for blue whale data.

@@

1. Remove noise and de-trend the acceleration signal with a 5th order Butterworth band-pass filter (killer whale: [1-25Hz], blue whale: [1-10Hz]) [R package `signal` @R-signal].
2. Enhance the IJK complex by differentiating acceleration using a 4th order Savitzky-Golay filter (R package `signal`). Differentiation exaggerates impulses like the J wave.
3. Further enhance the peaks by calculating the Shannon entropy ($H_i=-\sum_{k} |a_{ik}| \times ln(|a_{ik}|)$, where $k$ is the acceleration axis). Additionally, the Shannon entropy is strictly positive, which facilitates peak detection. In the 1d case, $k$ is surge (anterior-posterior acceleration) only.
4. Remove noise by applying a triangular moving average smoother.
5. Extract peaks and heuristically remove noisy peaks (see supplemental).

@@

We tested whether the 3d BCG was more robust than 1d BCG in field data by comparing the signal-to-noise ratios. For both BCGs, we calculated the power spectral density [R package `psd`; @Barbour2014]. Previously recorded blue whale apneic heart rate was 4-8 bpm [@goldbogenExtremeBradycardiaTachycardia2019], so we quantified _signal_ as the integration of the power spectral density curve from 4-8 bpm and _noise_ as the integrated remainder, up to 60 bpm.

We also tested whether BCG-derived instantaneous heart rates exhibited bradycardia relaxation over the course of dives, consistent with diving physiology patterns in marine mammals [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014]. We assigned dive start and end times when the whale swam deeper than 2 m, retaining dives that exceeded 10 m depth and 5 minutes duration. Dive times were normalized from 0 (start of dive) to 1 (end of dive). We regressed instantaneous heart rate against normalized dive time using robust Theil-Sen regression (to account for heteroscedascity) [R package `RobustLinearReg`; @R-RobustLinearReg; @Sen-1968; @theilRankInvariantMethodLinear1992] and tested whether the slope was greater than 0.

@@

The data and code used in this analysis were packaged as a research compendium [R package `rrtools`; @marwickPackagingDataAnalytical2018; @rrtools2019]. The research compendium was written as an R package so other researchers can read, run, and modify the methods described here.

@@

The ECG and BCG yielded nearly identical heart rate estimations (Fig. \@ref(fig:oo-bcg-ecg)). We collected 14 s of simultaneous ECG and BCG data during a motionless, submerged breath hold. BCG-derived instantaneous heart rates were within `r sprintf("%0.1f%% \u00B1 %0.1f%%", mean_bpm_diff * 100, sd_bpm_diff * 100)` of the ECG-derived rates (mean &#177; standard deviation). Ordinary least squares regression of BCG heartrates on ECG heartrates yielded a slope of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[2], bpm_stderr[2])` and intercept of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[1], bpm_stderr[1])` (mean &#177; standard error), which were not significantly different from the hypothesized 1 and 0, respectively.

@@

```{r motionless}
bw_elg <- bw180905_53_elg %>% 
  mutate(
    depth_min = map2_dbl(start, stop, function(t1, t2) {
      bw180905_53_10hz[between(bw180905_53_10hz$dt, t1, t2), "depth"] %>% min()
    }),
    depth_max = map2_dbl(start, stop, function(t1, t2) {
      bw180905_53_10hz[between(bw180905_53_10hz$dt, t1, t2), "depth"] %>% max()
    })
  )
n_motionless <- nrow(bw_elg)
motionless_minutes <- as.numeric(sum(bw_elg$stop - bw_elg$start), unit = "mins")
```

@@

The 3d BCG (Fig. \@ref(fig:bw-bcg)) produced a more robust signal than the 1d BCG, which used only anterior-posterior acceleration. The signal-to-noise ratio was 2.00 for the 3d BCG, compared to 0.17 for the 1d BCG (Fig. \@ref(fig:bw-validation-plots)**A**).

@@

```{r brady-relax-model}
bw10hz <- bw180905_53_10hz %>%
  mutate(
    dive_id = split_dives(dt, depth,
                          surface = 2, min_depth = 10, min_dur = 5 * 60),
    dive_norm = normalize_dives(dive_id),
    rid_left = approx(bw_elg$start, 
                      bw_elg$region_id, 
                      dt, 
                      "constant")$y,
    rid_right = approx(bw_elg$stop, 
                       bw_elg$region_id, 
                       dt, 
                       "constant", 
                       yleft = 0)$y + 1
  )
bw_400hz$dive_norm <- approx(bw10hz$dt, bw10hz$dive_norm, bw_400hz$dt)$y

@@

3d BCG-derived heart rates exhibited a relaxation of bradycardia over the course of dives. Average heart rate increased from `r format(bpm_pred$bcg_bpm[1], digits = 2)` bpm at the start of dives to `r format(bpm_pred$bcg_bpm[nrow(bpm_pred)], digits = 2)` bpm at the end of dives (Theil-Sen regression, $p < 10^{-10}$) (Fig. \@ref(fig:bw-validation-plots)**B**).

@@

Here we presented a ballistocardiogram method for detecting resting apneic heartrate in cetaceans using accelerometers. We validated the method in a controlled setting with simultaneous ECG and in a field setting by confirming expected physiological patterns. As accelerometer tags have been deployed on many cetacean species for multiple decades, this method may be applied to mine existing datasets and better understand how heartrate scales with body size and other biological factors. Current IMU tag designs limit BCG analysis to motionless periods, but future dimple- or limpet-style tags could reduce acceleration noise, boost the signal-to-noise ratio, and make the method more widely applicable. Even as the field of physio-logging progresses with new hardware innovations, this method demonstrates that computational advances can still derive new insights from traditional sensors.

@@

* Everyone who helped collect and process the blue whale data.
* The Sea World trainers.
* Anna Krystalli, Ben Marwick, Karthik Ram, Nicholas Tierney, and other members of the open science R community for developing tools and educational resources that facilitate open science practices.
* Funding from Office of Naval Research N000141912455, Stanford Terman Fellowship
* M. F. Czapanskiy was supported by the Stanford University William R. and Sara Hart Kimball Fellowship and a Stanford Data Science Scholar Fellowship.

@@

```{r oo-bcg-ecg, fig.cap="The ECG (**A**) and BCG (**E**), recorded by the ECG and CATS IMU tags, respectively, produced nearly identical heart beat predictions for the killer whale. **B-D** display the intermediate steps in the BCG signal processing procedure. **B**: Anterior-posterior axis acceleration (surge) after band-pass filtering. Inset shows the IJK complex with surrounding H and L waves for the region bounded by the dashed box. **C**: Peaks enhanced after forward differencing. **D**: A strictly positive signal after calculating Shannon entropy. Y-axis labeling follows [@leePhysiologicalSignalMonitoring2016]; y-axis values excluded because filtering introduces magnitude distortion and only the relative shape of the signal is relevant to the analysis."}
t <- theme_classic() +
  theme(axis.title.x = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_blank())
t_yvoid <- t +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())

@@

# B: Filtered surge
ijk_range <- as.POSIXct("2021-08-16 10:09:31.4", tz = "US/Pacific") + c(0, 0.4)
ijk_complex <- oo_400hz %>% 
  filter(between(dt, ijk_range[1], ijk_range[2])) %>% 
  mutate(t = as.numeric(dt - min(dt), unit = "secs"))
ijk_labels <- slice(ijk_complex, c(41, 53, 67, 85, 110)) %>% 
  mutate(wave = c("H", "I", "J", "K", "L"))
pB_inset <- ggplot(ijk_complex, aes(dt, surge_filt)) + 
  geom_line() +
  geom_text(aes(label = wave),
            ijk_labels,
            vjust = c(-0.2, 1.2, 0.75, -0.6, -0.2),
            hjust = c(0.5, 0.5, -0.6, 0.5, 0.5),
            fontface = "bold",
            size = 3) +
  theme_void()

inset_range <- as.POSIXct("2021-08-16 10:09:29.9", tz = "US/Pacific") + c(0, 4)
pB <- ggplot(oo_400hz, aes(dt, surge_filt)) +
  geom_line(size = 0.2) +
  annotation_custom(grob = ggplotGrob(pB_inset),
                    xmin = inset_range[1], xmax = inset_range[2],
                    ymin = 0.25, ymax = 0.5) +
  annotate("rect",
           xmin = ijk_range[1], 
           xmax = ijk_range[2],
           ymin = min(ijk_complex$surge_filt) - 0.02, 
           ymax = max(ijk_complex$surge_filt) + 0.02,
           fill = NA,
           color = "grey50",
           lty = 3) +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  expand_limits(y = 0.5) +
  labs(y = bquote(f[HR])) +
  t_yvoid

# C: Surge diff
pC <- ggplot(oo_400hz, aes(dt, surge_diff)) +
  geom_line(size = 0.2) +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = "dBCG") +
  t_yvoid

# D: Shannon entropy
pD <- ggplot(oo_400hz, aes(dt, surge_se)) +
  geom_line(size = 0.2) +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = "seBCG") +
  t_yvoid

# E: Smoothed
pE <- ggplot(oo_400hz, aes(dt, surge_smooth)) +
  geom_line(size = 0.2) +
  geom_point(data = filter(oo_400hz, bcg_beat), color = "blue") +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = "Y") +
  t_yvoid

cowplot::plot_grid(pA, pB, pC, pD, pE,
                   align = "v", 
                   axis = "lr",
                   ncol = 1, 
                   rel_heights = c(1, 1.5, 0.75, 0.75, 1),
                   labels = "AUTO")
```

@@

```{r bw-bcg, fig.cap="Example of signal processing for 3d BCG during a motionless period in a blue whale dive. **A**: Band-pass filtered triaxial acceleration, where surge is along the anterior-posterior axis, sway is along the lateral axis, and heave is along the dorso-ventral axis. **B**: Differencing the filtered acceleration enhances peaks. **C**: Calculating the Shannon entropy combines information from all three axes and makes the signal strictly positive. **D**: Smoothing the Shannon entropy facilitates robust peak detection. Detected heart beats in blue. Y-axis labeling follows [@leePhysiologicalSignalMonitoring2016]; y-axis values were excluded because the filtering process introduces magnitude distortion and only the relative shape of the signal is relevant to the analysis."}
bw_bcg_plot <- function(rid) {
  bcg_sample <- filter(bw_400hz, region_id == rid)

  # A: Filtered surge
  pA <- bcg_sample %>% 
    rename(Surge = surge_filt, Sway = sway_filt, Heave = heave_filt) %>% 
    pivot_longer(Surge:Heave, names_to = "axis", values_to = "acc") %>% 
    mutate(axis = factor(axis, levels = c("Surge", "Sway", "Heave"))) %>% 
    ggplot(aes(dt, acc)) +
    geom_line(size = 0.2) +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    scale_y_continuous(n.breaks = 3) +
    facet_grid(rows = vars(axis)) +
    labs(y = bquote(f[HR])) +
    t_yvoid
